Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-11-22 paper/manuscript.tex

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{azimi2022relative} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}~\cite{daviddotson2020alchemistry}.
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper      
	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
	\item Cinnabar: Standardising alchemical free energy analysis. Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
    \url{https://github.com/michellab/freenrgworkflows}.
    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
	\end{itemize}
\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis but allow more flexibility and modification.
\end{itemize}

@@

\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, and and contributed to the discussion and review of many sections.\\

>>>>

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{azimi2022relative} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}~\cite{daviddotson2020alchemistry}.
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper      
	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
	\item \href{https://github.com/OpenFreeEnergy/cinnabar}{Cinnabar}: Standardising alchemical free energy analysis. Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
	\end{itemize}
\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis but allow more flexibility and modification.
\end{itemize}

@@

\textbf{EG} Wrote the updated Sec.~\ref{sec:theory} and Sec.~\ref{sec:multiple-chemical-states},  updated Sec.~\ref{sec:standardstate-restraints}, and contributed to the discussion and review of many sections.\\
